---
title: "CIA_Analysis"
author: "Nasim Bondar Sahebi"
date: "2024-02-13"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Load Libraries

```{r  load libraries}
library(SummarizedExperiment)
library(EnhancedVolcano)
```


## Data Loading and Preparation

#### Data Overview

  - **dataset name **:  CIA
  - **Patient samples **: 50 
  - **RNA sequencing (RNA-seq)**: 50 patients, 51627 genes.
  - **LOW **:  43 patients
  - **HIGH **:  7 patients
  
  
Load the Summarized Experiment data object, extract clinical information, log-transformed FPKM expression data, along with annotations. Prepare the gene expression data for analysis.

```{r summery load-data}

# Load your SE result and extract clinical data , expression data and annotation

# Load SE obj
se <- readRDS("~/BHK lab/Pancreas-cancer/output data/SE_CIA.rds")

# Extract Clinical data 
clin <- data.frame(colData(se)) #dim is 50 x 16

# Extract the expression data
expr_fpkm <- assays(se)[["gene_expression"]] #dim 51627 x 50

# Extract the annotation
annot <- data.frame(rowData(se))  #dim is 51627 x 15

# Display first few rows of the dataset
DT::datatable(round(expr_fpkm[1:8, 1:7], 3)) 

```

## Data Preparation 

Preparing expression data (RNA-seq) for analysis:

1. Restrict expression data to protein-coding genes.
2. Remove low/zero expression genes.

Total 17354 genes remain for downstream analyses. 

```{r Data Preparation }

# Step 1: Restrict to Protein-Coding Genes.
annot_proteincoding <- annot[annot$gene_type == "protein_coding",] # 19477 protein coding genes.
expr_fpkm<- expr_fpkm[rownames(expr_fpkm) %in% rownames(annot_proteincoding),]
dim(expr_fpkm)

# Step 2: Filter Low/Zero Expression Genes]

#data is log2(TPM+1)
r <- as.numeric(apply(expr_fpkm, 1, function(i) sum(round(i, 6) == round(log2(1), 6))))

# Get the indices of rows to remove
remove <- which(r > dim(expr_fpkm)[2] * 0.5) 

# Remove rows from the matrix
expr_fpkm <- expr_fpkm[-remove, ]   #dim is 17354 x 50

DT::datatable(round(expr_fpkm[1:8, 1:8], 3))
```


## Differential gene expression.

  - **Use limma approach in R **: To find the differentially expressed genes between LOW and HIGH.
  - **P-values are corrected to control false discovery rate (FDR)** using Benjamini-Hochberg approach 
  - **out put:** CIA_Limma_Fit.txt file
  
result: only one DE genes
  
```{r diff expression }

# Limma approach applied.
design <- model.matrix(~ clin$low_high)
fit <- lmFit(expr_fpkm, design)
fit <- eBayes(fit)

# Sort by Pvalue.
top.table <- topTable(fit, sort.by = "P", n = Inf)

# To see How many DE gene are there.
length(which(top.table$adj.P.Val < 0.05))  # only one DE genes.

# Preview
DT::datatable(data.frame(round(data.frame(top.table),3)))

# Save top table to .txt file.
top.table$Gene <- rownames(top.table)
write.table(top.table, file = "~/BHK lab/kevin Project/Cholangiocarcinoma/output data/Limma_Fit.txt", row.names = F, sep = "\t", quote = F)

```

