---
title: "CIA_Analysis"
author: "Nasim Bondar Sahebi"
date: "2024-02-13"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Load Libraries

```{r  load libraries}
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(knitr)
library(DT)
library(limma)
library(EnhancedVolcano)
library(fgsea)
library(msigdbr)
library(gage)
library(pathview)
```


## Data Loading and Preparation

#### Data Overview

  - **dataset name **:  CIA
  - **Patient samples **: 50 
  - **RNA sequencing (RNA-seq)**: 50 patients, 51627 genes.
  - **LOW **:  43 patients
  - **HIGH **:  7 patients
  
  
Load the Summarized Experiment data object, extract clinical information, log-transformed FPKM expression data, along with annotations. Prepare the gene expression data for analysis.

```{r summery load-data}

# Load your SE result and extract clinical data , expression data and annotation

# Load SE obj
se <- readRDS("~/BHK lab/Pancreas-cancer/output data/SE_CIA.rds")

# Extract Clinical data 
clin <- data.frame(colData(se)) #dim is 50 x 16

# Extract the expression data
expr_fpkm <- assays(se)[["gene_expression"]] #dim 51627 x 50

# Extract the annotation
annot <- data.frame(rowData(se))  #dim is 51627 x 15

# Display first few rows of the dataset
DT::datatable(round(expr_fpkm[1:8, 1:7], 3)) 

```

## Data Preparation 

Preparing expression data (RNA-seq) for analysis:

1. Restrict expression data to protein-coding genes.
2. Remove low/zero expression genes.

Total 17354 genes remain for downstream analyses. 

```{r Data Preparation }

# Step 1: Restrict to Protein-Coding Genes.
annot_proteincoding <- annot[annot$gene_type == "protein_coding",] # 19477 protein coding genes.
expr_fpkm<- expr_fpkm[rownames(expr_fpkm) %in% rownames(annot_proteincoding),]
dim(expr_fpkm)

# Step 2: Filter Low/Zero Expression Genes]

#data is log2(TPM+1)
r <- as.numeric(apply(expr_fpkm, 1, function(i) sum(round(i, 6) == round(log2(1), 6))))

# Get the indices of rows to remove
remove <- which(r > dim(expr_fpkm)[2] * 0.5) 

# Remove rows from the matrix
expr_fpkm <- expr_fpkm[-remove, ]   #dim is 17354 x 50

DT::datatable(round(expr_fpkm[1:8, 1:8], 3))
```


## Differential gene expression.

  - **Use limma approach in R **: To find the differentially expressed genes between LOW and HIGH.
  - **P-values are corrected to control false discovery rate (FDR)** using Benjamini-Hochberg approach 
  - **out put:** CIA_Limma_Fit.txt file
  
result: only one DE genes
  
```{r diff expression }

# Limma approach applied.
design <- model.matrix(~ clin$low_high)
fit <- lmFit(expr_fpkm, design)
fit <- eBayes(fit)

# Sort by Pvalue.
top.table <- topTable(fit, sort.by = "P", n = Inf)

# To see How many DE gene are there.
length(which(top.table$adj.P.Val < 0.05))  # only one DE genes.

# Preview
DT::datatable(data.frame(round(data.frame(top.table),3)))

# Save top table to .txt file.
top.table$Gene <- rownames(top.table)
write.table(top.table, file = "~/BHK lab/Pancreas-cancer/output data/CIA_Limma_Fit.txt", row.names = F, sep = "\t", quote = F)

```

- **Volcano Plot:** based on P value and FDR

```{r  Volcano plot libraries,fig.width=10, fig.height=10}


# 1. Volcano plot.

# 1.1 Volcano Plot based on P value
EnhancedVolcano(top.table,
    lab = rownames(top.table),
    x = 'logFC',
    y = 'P.Value',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    xlim = c(-5, 5),
    ylim = c(0, -log10(10e-4)),
    title= 'Volcano Plot based on P value'
    )

# 1.2 Volcano Plot based on FDR 
# adjusted p-values as red 
EnhancedVolcano(top.table,
    lab = rownames(top.table),
    x = 'logFC',
    y = 'adj.P.Val',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    xlim = c(-5, 5),
    ylim = c(0, -log10(10e-4)),
    legendLabels=c('Not sig.','Log FC','adj p-value',
      'adj p-value & Log FC'),
    title= 'Volcano Plot based on FDR',)

```

## MSigDB Pathway analysis

- To Identify the associated pathways or genes sets.
- Downloaded hallmark(H) , KEGG gene sets from [ MSigDB website](https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp#H).
- Conduct GSEA with the fgsea package using "top.table" from step 2.

## Hallmark Geneset.

```{r  Create Rank and Bar pot}

# Load human hallmark gene sets from MSigDB
pathwaysDF <- msigdbr("Homo sapiens", category="H")

# Create a list of gene symbols for each pathway
pathwaysH <- split(as.character(pathwaysDF$gene_symbol), pathwaysDF$gs_name)

# 1. Create ranks

# Create 'ranks' vector from 'top.table' with logFC as values and gene names as names.
ranks <- top.table$logFC
names(ranks) <- top.table$Gene
ranks <- sort(ranks, decreasing = T)

# Preview the data table of ranks.
DT::datatable(data.frame(ranks))

# 2. Bar Plot
# Display the ranked fold changes.
barplot(ranks)
```


```{r  Pathway analysis, fig.width=10, fig.height=10 }
# 3. Conduct analysis
# Run fgsea with 'pathwaysH' and 'ranks'.
fgseaRes <- fgsea(pathwaysH, ranks)
fgseaRes <- na.omit(fgseaRes)

fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES)) #order by NES

# Combine genes in leadingEdge with semicolons
fgseaResTidy$leadingEdge <- sapply(fgseaResTidy$leadingEdge, paste, collapse = ";")
fgseaResTidy <- fgseaResTidy[order(fgseaResTidy$padj), ]

# Show in a Tidy table.
head(fgseaResTidy)

# Save fgseaRes as txt file.
write.table(fgseaResTidy, "~/BHK lab/Pancreas-cancer/output data/CIA_fgseaRes_HRank.txt", sep = "\t", row.names = FALSE, quote = FALSE)

# Mark "significant" if adjusted p-value <= 0.05, else "non-significant".
fgseaResTidy$adjPvalue <- ifelse(fgseaResTidy$padj <= 0.05, "significant", "non-significant")

# Define colors for significance: grey for non-significant and red for significant results.
cols <- c("non-significant" = "grey", "significant" = "red")

# Using ggplot: reorder pathways by NES, fill bars based on significance, flip coordinates for readability.
ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES, fill = adjPvalue)) + geom_col() +
    scale_fill_manual(values = cols) + coord_flip() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(x = "Pathway", y = "Normalized Enrichment Score", title = "Hallmark pathways NES
         Enrichment Score from GSEA")

# 4. GSEA Results Table Plot.
# Plot GSEA table for significant pathways (padj < 0.05) using a GSEA parameter of 0.5.
plotGseaTable(pathwaysH[fgseaRes$pathway[fgseaRes$padj < 0.05]], ranks, fgseaRes, gseaParam=0.5)
```

```{r  Enrichment score plot }
# 5.  Enrichment score plot.

# For significant pathways # 18 pathways.

# Store the list of significant pathways (padj < 0.05).
significant_pathways <- fgseaRes$pathway[fgseaRes$padj < 0.05]

# Use a for loop to iterate and plot enrichment for each significant pathway
for (i in significant_pathways) {
  enrichment_plot <- plotEnrichment(pathwaysH[[i]], ranks)
  enrichment_plot <- enrichment_plot + ggtitle(paste("Enrichment Plot for Pathway:",i))
  print(enrichment_plot)
}

# Plot significant pathways as points with a gradient color based on p-value and size based on a 'size' variable.
ggplot(fgseaResTidy[fgseaResTidy$padj < 0.05, ], aes(y=reorder(pathway, size),x= NES)) + geom_point(aes(color=padj,size=size)) + 
  scale_color_gradientn(colours = c("red","blue")) + 
  labs(x='Normalized Enrichment Score', y=NULL ) + 
  theme( axis.title = element_text(face='bold'), axis.text = element_text (face='bold') )

```



## KEGG Pathway Analysis 

To do KEGG pathway analysis, both KEGG_MEDICUS and KEGG_LEGACY are downloaded. The cutoff of 0.05 is applied for visualization, while the text files include the results across all gene sets.

## KEGG_MEDICUS 

619 canonical pathways gene sets derived from the KEGG MEDICUS pathway database.

```{r  KEGG plot,fig.width= 13 }

# Load KEGG_MEDICUS pathways.
pathwaysKEG <- gmtPathways("~/BHK lab/Pancreas-cancer/GSEA/c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")

# Perform fgsea with KEGG pathways
fgseaRes_kegg <- fgsea(pathwaysKEG, ranks)

# Organize and sort KEGG results by NES
fgseaResTidy_kegg <- fgseaRes_kegg %>%
  as_tibble() %>%
  arrange(desc(NES))

# Combine genes in leadingEdge with semicolons
fgseaResTidy_kegg$leadingEdge <- sapply(fgseaResTidy_kegg$leadingEdge, paste, collapse = ";")

# Show in a Tidy table of keg.
head(fgseaResTidy_kegg)

# Save fgseaRes_keg as txt file.
write.table(fgseaResTidy_kegg, "~/BHK lab/Pancreas-cancer/output data/CIA_fgseaRes_kegRank.txt", sep = "\t", row.names = FALSE, quote = FALSE)
 

```

```{r KEGG plot ,fig.width= 9 , fig.height= 6 }

# Mark "significant" if adjusted p-value <= 0.05, else "non-significant".
fgseaResTidy_kegg$adjPvalue <- ifelse(fgseaResTidy_kegg$padj <= 0.05, "significant", "non-significant")

# Create a ggplot of significant pathways, with bars colored based on adjusted p-value significance.
ggplot(fgseaResTidy_kegg[fgseaResTidy_kegg$padj < 0.05, ], aes(reorder(pathway, NES), NES, fill = adjPvalue)) + 
  geom_col() + 
  scale_fill_manual(values = cols) + 
  coord_flip() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Pathway", y = "Normalized Enrichment Score", title = "KEGG pathways NES Enrichment Score from GSEA")


# Plot significant pathways as points with a gradient color based on p-value and size based on a 'size' variable.
ggplot(fgseaResTidy_kegg[fgseaResTidy_kegg$padj < 0.05, ], aes(y=reorder(pathway, size),x= NES)) + geom_point(aes(color=padj,size=size)) + 
  scale_color_gradientn(colours = c("red","blue")) + 
  labs(x='Normalized Enrichment Score', y=NULL ) + 
  theme( axis.title = element_text(face='bold'), axis.text = element_text (face='bold') )

```
## KEGG_LEGACY

186 canonical pathways gene sets derived from the KEGG pathway database. These are considered Legacy gene sets since the introduction of the gene sets based on the more recent KEGG MEDICUS data. 

```{r  KEGG legacy,fig.width= 13 }

# Load KEGG_MEDICUS pathways.
pathwaysKEG_LEG <- gmtPathways("~/BHK lab/Pancreas-cancer/GSEA/c2.cp.kegg_legacy.v2023.2.Hs.symbols.gmt")

# Perform fgsea with KEGG pathways
fgseaRes_kegg <- fgsea(pathwaysKEG, ranks)

# Organize and sort KEGG results by NES
fgseaResTidy_kegg <- fgseaRes_kegg %>%
  as_tibble() %>%
  arrange(desc(NES))

# Combine genes in leadingEdge with semicolons
fgseaResTidy_kegg$leadingEdge <- sapply(fgseaResTidy_kegg$leadingEdge, paste, collapse = ";")

# Show in a Tidy table of keg.
head(fgseaResTidy_kegg)

# Save fgseaRes_keg as txt file.
write.table(fgseaResTidy_kegg, "~/BHK lab/Pancreas-cancer/output data/CIA_fgseaRes_keggLEGRank.txt", sep = "\t", row.names = FALSE, quote = FALSE)

```


```{r KEGG legacy plot,fig.width= 9 , fig.height= 6 }

# Mark "significant" if adjusted p-value <= 0.05, else "non-significant".
fgseaResTidy_kegg$adjPvalue <- ifelse(fgseaResTidy_kegg$padj <= 0.05, "significant", "non-significant")


# Create a ggplot of significant pathways, with bars colored based on adjusted p-value significance.
ggplot(fgseaResTidy_kegg[fgseaResTidy_kegg$padj < 0.05, ], aes(reorder(pathway, NES), NES, fill = adjPvalue)) + 
  geom_col() + 
  scale_fill_manual(values = cols) + 
  coord_flip() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Pathway", y = "Normalized Enrichment Score", title = "KEGG (legacy) pathways NES Enrichment Score from GSEA")


# Plot significant pathways as points with a gradient color based on p-value and size based on a 'size' variable.
ggplot(fgseaResTidy_kegg[fgseaResTidy_kegg$padj < 0.05, ], aes(y=reorder(pathway, size),x= NES)) + geom_point(aes(color=padj,size=size)) + 
  scale_color_gradientn(colours = c("red","blue")) + 
  labs(x='Normalized Enrichment Score', y=NULL ) + 
  theme( axis.title = element_text(face='bold'), axis.text = element_text (face='bold') )

```

## Over-Representation Analysis analysis (ORA)

To identifies whether certain gene sets are over-represented in a target gene group.

```{r  ORA legacy }

# Set a vector of differentially expressed genes for ORA
deGenes <- rownames(top.table)[top.table$adj.P.Val < 0.05]  # 1 genes

# Defining the gene universe as all genes considered in the analysis
geneUniverse <- rownames(expr_fpkm)

## Performing ORA

# ORA for Hallmark pathways
oraHallmark <- fora(pathways = pathwaysH, genes = deGenes, universe = geneUniverse, minSize = 15, maxSize = 500)
oraHallmark$Pathway_Set <- "Hallmark"
oraHallmark$Significance <- ifelse(oraHallmark$padj < 0.05, "Significant", "Not Significant")

# ORA for KEGG_MEDICUS pathways
oraKeggMed <- fora(pathways = pathwaysKEG, genes = deGenes, universe = geneUniverse, minSize = 15, maxSize = 500)
oraKeggMed$Pathway_Set <- "KEGG_MEDICUS"
oraKeggMed$Significance <- ifelse(oraKeggMed$padj < 0.05, "Significant", "Not Significant")

# ORA for KEGG_LEGACY pathways
oraKeggLeg <- fora(pathways = pathwaysKEG_LEG, genes = deGenes, universe = geneUniverse, minSize = 15, maxSize = 500)
oraKeggLeg$Pathway_Set <- "KEGG_LEGACY"
oraKeggLeg$Significance <- ifelse(oraKeggLeg$padj < 0.05, "Significant", "Not Significant")

# Display the top results for each pathway set
# DT::datatable(head(oraHallmark, 
#               options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 4, autoWidth = TRUE))
DT::datatable(oraHallmark)
DT::datatable(oraKeggMed)
DT::datatable(oraKeggLeg)


```

### Visulize ORA 
- Heatmap 

```{r OraHallmark,fig.width= 14 , fig.height=10}

# oraHallmark
# Create an enrichment heatmap
ggplot(oraHallmark, aes(x = pathway, y = reorder(pathway, padj), fill = -log10(padj))) +
  geom_tile() +
  scale_fill_gradientn(colors = c("blue", "yellow", "red"), na.value = "grey50", trans = "reverse") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(fill = "-log10(p-adj)", x = "Pathway", y = "") +
  ggtitle("Enrichment Heatmap of Top Pathways")


```

```{r ORA keggleg ,fig.width= 14 , fig.height=10}

#ORA Kegg

# Create an enrichment heatmap
ggplot(oraKeggLeg[1:50, 1:8], aes(x = pathway, y = reorder(pathway, padj), fill = -log10(padj))) +
  geom_tile() +
  scale_fill_gradientn(colors = c("blue", "yellow", "red"), na.value = "grey50", trans = "reverse") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(fill = "-log10(p-adj)", x = "Pathway", y = "") +
  ggtitle("Enrichment Heatmap of Top Pathways")


```


```{r orakeggmed ,fig.width= 20 , fig.height=17}

# Create an enrichment heatmap
ggplot(oraKeggMed[1:50, 1:8], aes(x = pathway, y = reorder(pathway, padj), fill = -log10(padj))) +
  geom_tile() +
  scale_fill_gradientn(colors = c("blue", "yellow", "red"), na.value = "grey50", trans = "reverse") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(fill = "-log10(p-adj)", x = "Pathway", y = "") +
  ggtitle("Enrichment Heatmap of Top Pathways")

```

