---
title: "TCGA_Analysis"
author: "Nasim Bondar Sahebi"
date: "2024-02-13"
output: html_document
pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load Libraries

```{r  load libraries}
library(SummarizedExperiment)
library(limma)
library(EnhancedVolcano)
```


## Data Loading and Preparation

#### Data Overview

  - **Paper name **:  Integrated Genomic Characterization of Pancreatic Ductal Adenocarcinoma
  - ** PMID **: [28810144](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5964983/)
  - **Patient samples **: 120 
  - **RNA sequencing (RNA-seq)**: 120 patients, 18416 genes.
  - **LOW **: 82 patients
  - **HIGHT **:  38 patients
  
  
Load the Summarized Experiment data object, extract clinical information, log-transformed FPKM expression data, along with annotations. Prepare the gene expression data for analysis.

```{r summery load-data}

# Load your SE result and extract clinical data , expression data and annotation

# Load SE obj
se <- readRDS("~/BHK lab/Pancreas-cancer/output data/SE_TCGA.rds")

# Extract Clinical data 
clin <- data.frame(colData(se)) #dim is 120 x 24

# Extract the expression data
expr_fpkm <- assays(se)[["gene_expression"]] #dim is 18416 x 120

# Extract the annotation
annot <- data.frame(rowData(se))  #dim is 18416 x 15

# Display first few rows of the dataset
DT::datatable(round(expr_fpkm[1:8, 1:5], 3)) 

```

## Data Preparation 

Preparing expression data (RNA-seq) for analysis:

1. Restrict expression data to protein-coding genes.
2. Remove low/zero expression genes.

Total 15,884 genes remain for downstream analyses. 

```{r Data Preparation }

# Step 1: Restrict to Protein-Coding Genes.
annot_proteincoding <- annot[annot$gene_type == "protein_coding",] # 17586 protein coding genes.
expr_fpkm <- expr_fpkm[rownames(expr_fpkm) %in% rownames(annot_proteincoding),]
dim(expr_fpkm)

# Step 2: Filter Low/Zero Expression Genes]

# Data is log2(FPKM+1)
r <- as.numeric(apply(expr_fpkm, 1, function(i) sum(round(i, 6) == round(log2(1), 6))))

# Get the indices of rows to remove
remove <- which(r > dim(expr_fpkm)[2] * 0.5) 

# Remove rows from the matrix
expr_fpkm <- expr_fpkm[-remove, ]   #dim is 15884 x 120

DT::datatable(round(expr_fpkm[1:8, 1:5], 3)) 
```


## Differential gene expression.

  - **Use limma approach in R **: To find the differentially expressed genes between LOW and HIGH.
  - **P-values are corrected to control false discovery rate (FDR)** using Benjamini-Hochberg approach 
  - **out put:** TCGCA_Limma_Fit.txt file
  

```{r diff expression }

# Limma approach applied.
design <- model.matrix(~ clin$low_high)

# Fit the model
fit <- lmFit(expr_fpkm, design)

# Calculate the t-statistics 
fit <- eBayes(fit)

# Sort by Pvalue.
top.table <- topTable(fit, sort.by = "P", n = Inf)

# To see How many DE gene are there.
length(which(top.table$adj.P.Val < 0.05))  # no DE genes.

# Preview
DT::datatable(round(data.frame(top.table),3))

# Save top table to .txt file.
top.table$Gene <- rownames(top.table)
write.table(top.table, file = "~/BHK lab/Pancreas-cancer/output data/TCGCA_Limma_Fit.txt", row.names = F, sep = "\t", quote = F)

```

## Volcano plot

 - **Volcano Plot:** based on P value and FDR

```{r  Volcano plot libraries,fig.width=10, fig.height=10}


# 1. Volcano plot.

# 1.1 Volcano Plot based on P value
EnhancedVolcano(top.table,
    lab = rownames(top.table),
    x = 'logFC',
    y = 'P.Value',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    xlim = c(-5, 5),
    ylim = c(0, -log10(10e-4)),
    title= 'Volcano Plot based on P value'
    )

# 1.2 Volcano Plot based on FDR 
# adjusted p-values as red 
EnhancedVolcano(top.table,
    lab = rownames(top.table),
    x = 'logFC',
    y = 'adj.P.Val',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    xlim = c(-5, 5),
    ylim = c(0, -log10(10e-4)),
    legendLabels=c('Not sig.','Log FC','adj p-value',
      'adj p-value & Log FC'),
    title= 'Volcano Plot based on FDR',)

```

